---
title: "Data Carpentry Workshop 2022 Code"
author: "Erin Witkop"
date: "5/13/2022"
output: github_document

---
## DAY 1

### 9:00 am Introduction 

Welcome to the second data carpentry style workshop that the Bioinformatics group has hosted. My name is Erin Witkop, I am a Bioinformatics Postdoctoral Research Associate in Peter Linsley's lab and I started last October. I started coding in R as a undergrad and now have about 8 years of experience and about three years ago I became a certified Data Carpentry instructor, which means I have taken several of their courses, taught several prior courses, and have taken a course in the pedagogical style of the Carpentries. I really enjoy teaching and working with data and I'm excited to be teaching this course along with several fellow intructors Hannah DeBerg and Matt Dufort who are going to step in toward the end of the course.

This course is designed to start out with the assumption that you have no prior experience coding at all in any language and the curriculum is designed to build on itself throughout the course. Our main goals for the course are that in the future when you have your own data you will be able to load it into R, calculate some standard statisitcs, and do some exploratory plotting that you can share with your colleagues. 

We are basing the curriculum and pedagogical style of this course off of the Data Carpentries style, specifically the visusalizing ecology data course. 

For information regarding the schedule, helpful Data Carpentry resources, links to the code we will generate, and links to the collaborative etherpad document, please see our github repository [link](https://github.com/BenaroyaResearch/Data_Carpentry_Workshop_2022)

We are additionally going to use a collaborative etherpad document at several points during the course. The link for the etherpad is [here](https://pad.riseup.net/p/BRI_Data_Carpentry_2022).

Finally, I wanted our helpers for the workshop today:    . Every day we will have atleast two helpers who have experience with R Their role is to help troubleshoot any individual problems or error codes you may have that you are not able to figure out. Whenever you need help just raise up your red sticky note and a helper will come over and see if they can help assist you. At some points in the course I'll check in regarding the pace of the course and ask you to put up a green sticky note if you feel good, or red if I need to slow down and go slower.

I do want to point out that there are a lot of things to know about R, but the goal of this course is to help you build confidence and know that you find answers to your own questions.

### 9:10 am - 9:45am: Intro to R 

- What is R? What is RStudio?
  - The term “R” is used to refer to both the programming language and the software that          interprets the scripts written using it.

- Why learn R?
  - R is great for reproducibility, R can integrate new and existing tools, R is interdiscipinary and extensible, its free, is open source, R produces high quality graphics, large helpful community 
    - Stack overflow, R studio community
    
#### Knowing your way around R and R studio 

Let’s start by learning about RStudio, which is an Integrated Development Environment (IDE) for working with R.

The RStudio IDE open-source product is free under the Affero General Public License (AGPL) v3. The RStudio IDE is also available with a commercial license and priority email support from RStudio, PBC.

We will use RStudio IDE to write code, navigate the files on our computer, inspect the variables we are going to create, and visualize the plots we will generate. RStudio can also be used for other things (e.g., version control, developing packages, writing Shiny apps) that we will not cover during the workshop.

RStudio is divided into 4 “panes”:

  - The Source for your scripts and documents (top-left, in the default layout)
  - Your Environment/History (top-right) which shows all the objects in your working space (Environment) and your command history (History)
  - Your Files/Plots/Packages/Help/Viewer (bottom-right)
  - The R Console (bottom-left)

The placement of these panes and their content can be customized (see menu, Tools -> Global Options -> Pane Layout). For ease of use, settings such as background color, font color, font size, and zoom level can also be adjusted in this menu (Global Options -> Appearance).

One of the advantages of using RStudio is that all the information you need to write code is available in a single window. Additionally, with many shortcuts, autocompletion, and highlighting for the major file types you use while developing in R, RStudio will make typing easier and less error-prone.

#### Getting set up 

It is good practice to keep a set of related data, analyses, and text self-contained in a single folder, called the working directory. All of the scripts within this folder can then use relative paths to files that indicate where inside the project a file is located (as opposed to absolute paths, which point to where a file is on a specific computer). Working this way allows you to move your project around on your computer and share it with others without worrying about whether or not the underlying scripts will still work.

RStudio provides a helpful set of tools to do this through its “Projects” interface, which not only creates a working directory for you, but also remembers its location (allowing you to quickly navigate to it) and optionally preserves custom settings and (re-)open files to assist resume work after a break. Go through the steps for creating an “R Project” for this tutorial below.

  1. Start RStudio.
  2. Under the `File` menu, click on `New Project`. Choose `New Directory`, then `New Project`.
  3. Enter a name for this new folder (or “directory”), and choose a convenient location for it. This will be your working directory for the rest of the course (e.g., ~/data-carpentry).
  4. Click on Create Project.

A workspace is your current working environment in R which includes any user-defined object. By default, all of these objects will be saved, and automatically loaded, when you reopen your project.

#### Organizing your working directory

Using a consistent folder structure across your projects will help keep things organized, and will help you to find/file things in the future. This can be especially helpful when you have multiple projects. In general, you may create directories (folders) for scripts, data, and documents.

Let's create a few new folders to hold input and output for us in our repository
  - `/raw_data/`
  - `/scripts/`
  - `/figures/`
  - `/results/`

#### The working directory

The working directory is an important concept to understand. It is the place from where R will be looking for and saving the files. When you write code for your project, it should refer to files in relation to the root of your working directory and only need files within this structure.

RStudio assists you in this regard and sets the working directory automatically to the directory where you have placed your project in. If you need to check it, you can use `getwd()`. If for some reason your working directory is not what it should be you can use `setwd("/path/to/working/directory")` to reset your working directory.

#### Interacting with R

What we are writing is *code* or the instructions to tell R how to run. We call the instructions *commands* and then we execute the commands by running the code.

You can either run code directly through the console and press `Enter`, or by writing your commands in a script and running them using the `Run` button or a keyboard shortcut. For example, I run my commands in R on a mac using COMMAND+ENTER.

For reproducibility its best to write your code down in R and using the number (#) sign to include comments. Any line or part of a line starting with a number sign will be ignored by R. It's really important to comment your code, both for yourself and for sharing what you have done with others. 

#### Seeking help

You can search functions by typing `?mean()` in the console or also through the help tab in R studio. Looking up commands is really useful for understanding the arguments that each one takes and how to use it properly! 

#### Automatic Code Completion 

When you start typing code in R you can its tab completion to finish a command or object name that you were typing by hitting the tab button.

#### Dealing with Error Messages

Don’t get discouraged if your code doesn’t run immediately! Error messages are common when programming, and fixing errors is part of any programmer’s daily work. Often, the problem is a small typo in a variable name or a missing parenthesis. Watch for the red x’s next to your code in RStudio. These may provide helpful hints about the source of the problem.

RStudio shows a red x next to a line of code that R doesn’t understand.
RStudio shows a red x next to a line of code that R doesn’t understand.

If you can’t fix an error yourself, start by googling it. Some error messages are too generic to diagnose a problem (e.g. “subscript out of bounds”). In that case it might help to include the name of the function or package you’re using in your query.

### 9:45 - 10:30 am: Introduction to R 

#### Creating Objects in R 

You can get output from R simply by typing math in the console:

```{r math}
3 + 5
12 / 7
```

However to do things with this data we need to save it as an *object* by assigning the value to a name using the assignment operator `<-`

In RStudio, typing Alt + - (push Alt at the same time as the - key) will write <- in a single keystroke in a PC, while typing Option + - (push Option at the same time as the - key) does the same in a Mac.

```{r object}
age <- 50
```

R doesn't automatically print the output after you assign data to an object, so to view it you need to print the name of the object and run it

```{r object_print}
age
```

Objects can have almost any name you want, they just can't start with a number, have any spaces in them, or be the same name as one of the functions in R. R is case sensitive, and being consistent in your naming is helpful.  

For those of you familiar with any other coding languages, in R and `object` is the same thing as a `variable`. They are interchangable. 

Now that we have saved our data data as an object we can do math with it!

```{r age_math}
age * 2

# And we can save this output to a new object
age_double <- age * 2
```

You can also change the value by assigning a new one

```{r age_math_rename}
age <- 60
# Note this rewrites the previous data, but does not affect our doubled age object!

age_double

# So if you ever change a bit of code upstream, you need to rerun all the other commands that code depended on in order to update your data 
```

Let's do a quick challenge - put the answers to the following in the etherpad:

```{r challenge_1}
time_d <- 180
total_dosage_mg <- 500

# What is the dosage per day (mg)?
time_d/total_dosage_mg 

```

Now it's time to save your code! 

#### Functions and their arguments 

Functions are canned scripts that automate certain tasks, and all functions take arguments, and we can look these up using the help function! 

```{r sqrt}

age_sqrt <- sqrt(50)
age_sqrt

```

#### Vectors and Data Types 

A vector is the most common and basic data type in R, and is the workhorse of R. A vector is composed by a series of values, which can be either numbers or characters. We can assign a series of values to a vector using the c() function. `c` stands for concatenate. However, all the elements in the vector have to be the same type of data.

```{r age_c}

# They can contain numbers
age_d <- c(50, 100, 150, 200)
age_d 

# Or characters
treatments <- c("saline","tetramer","IL-13")
treatments
```

Several functions allow you to inspect a vector. 

``` {r vector_inspect}
# To view the number of elements
length(treatments)

# To view the type of data 
class(age_d) 
class(treatments)

# view the structure 
str(treatments)
str(age_d)

```

There are several other important data structures in R that we will discuss as we move through the course, including `list`, `matrix`, data frames (`data.frame`), factors (`factor`) and arrays (`arrays`).

##### Subsetting Vectors 

To extract a value from a vector you need to provide and index in square brackets.

```{r index_vector}

drug <- c("aspirin","tylenol","sudafed")

# to extract the third and second elements
drug[c(3,2)]

```

##### Conditional Subsetting

```{r condition}
weight_g <- c(25,30,35,40)

# Can ask a logical question
weight_g > 20
weight_g < 30

# you can also select values that meet a condition
weight_g[weight_g < 40]

# Or combine multiple tests with AND `&` and OR `|`
weight_g[weight_g <=30 | weight_g == 40]
```

You can also search for a list of strings in a vector using the `%in%` function 

```{r search}
months <- c("Jan","Feb","Mar","Apr")

# Use logical vector to find matches between lists
months[months %in% c("Jan","Mar","July")]

```

```{r challenge_2}

# can you find values between 100 and 200 in this list?
weight_lb <- c(55,95,105,187, 201)

```

##### Missing Data

Missing data in vectors is returned as `NA`. In some function you can add `na.rm = TRUE` to remove the `NA`s. 

```{r missing}
heights <- c(2,4,4,NA,6)
mean(heights) # why didn't this work?
mean(heights, na.rm = TRUE)

# You can also select for all values that are not NAs
heights[!is.na(heights)]

# You can also omit NAs
heights <- na.omit(heights)

```

```{r challenge_3}

# Can you return this data frame with no NAs?

weight_g <- c(10,15,16, NA,25, 30, NA)

```

#### BREAK: 10:30 - 10:45 ####

#### 10:45-11:15: Starting with Data in R: Dataframes ####

##### Downloading the Data

Now it's time to download a subset version of the COVID-19 data. The data is saved in a csv file and is saved on our Github landing page. https://github.com/BenaroyaResearch/Data_Carpentry_Workshop_2022/blob/main/WORKSHOP_DATA/Bolouri_2021_subset.csv

Let's open the csv table in EXCEL and view it. We have patient metadata, CYTOF data, and initial CBC counts for different cell populations. The data is in different formats also. 

Download the data to your raw_data folder 

```{r read_csv}
library(tidyverse)
setwd("/Users/ewitkop/Library/CloudStorage/Box-Box/EW_Bioinformatics_Postdoc_Research/ADMINISTRATIVE/Data_Carpentry_Workshop_June_2022/WORKSHOP_CODE/Data_Carpentry_Workshop_2022/")
data <- read_csv("raw_data/Bolouri_2021_subset.csv")

```

Note how read csv looks at the data type of each column. 

##### Inspecting data frames

```{r investigate}

# To view whole dataset
print(data,100)

# To see the first few lines
head(data)
head(data, 100)

# To see last few lines
tail(data)

# To open in R studio
View(data)

# additional helpful commands
nrow(data)
ncol(data)
rownames(data)
summary(data)

```

What we just loaded was a dataframe, the typical data structure used in R for tabular data while we will use for statistics and plotting.

In a dataframe columns are vectors of the same length and same type of data. 

```{r challenge_4}
# Based on the output of str(data), what is the class of our data object? How many rows and columns are there? 

str(data)
nrow(data) # 299
ncol(data) # 26
```

##### Indexing and subsetting data frames

To extract data from a data frame we need to use its coordinates, which are always row number followed by column number.

```{r index }
# get first row and column
data[1,1]

# You can save this by assigning it to a variable
data_subset <- data[1,1]

# get only 5th columns
data[,5]

# select group of rows and columns
data[1:2,3:6]

# You can also exclude columns using the minus sign
data[,-1]

# To get only the head of the dataframe you can do
data[-(7:nrow(data)),]

# You can also subset by using the name of the column in quotation marks
data[, "Sample.ID"]

# Finally you can also use the $ to get a column
data$Sample.ID

```

Challenge time!
```{r challenge_5}
#Create a data.frame (surveys_200) containing only the data in row 200 of the surveys dataset.

# Pull out the last row of the data frame and save it
```

Let's take five minutes for this challenge so you can have a breather before starting the final piece of our first day! 

#### 11:15-12:00pm: Starting with Data in R Cont: 

##### Factors 

Several of our data columns contain character data that are categorical variables. `Factor` is a special class used for working with categorical data. Factors only contain a list of pre-defined levels. Though factors may often look like character vectors, they are actually treated as integer vectors, so you need to be careful when treating them as strings.

```{r factors}

# Use the factor function to convert column levels to factors 
data$sex <- factor(data$sex) 

# check the conversion worked with summary and class
summary(data$sex)
class(data$sex)

# Note that factors are always assigned in alphabetical order. Check the levels of a factor with levels
levels(data$sex) # see that female comes before male

# Get the number of levels with n level 
nlevels(data$sex)
  
```

Sometimes you may want your levels in specific order.

```{r factor_reorder}
data$sex <- factor(data$sex, levels = c("male","female"))
levels(data$sex)

```

Challenge
```{r challenge_6}
# Convert the HigestCare column to a factor and put the levels in this order: Neg, Floor, CCU, Outpatient

levels(as.factor(data$HighestCare)) # "CCU"        "Floor"      "Neg"        "Outpatient"
data$HighestCare <- factor(data$HighestCare, levels = c("Neg","Floor","CCU","Outpatient"))
```

##### Converting factors

To convert a factor to character vector you can use `as.character(x)`

```{r convert_factor}
as.character(data$sex)
```

##### Renaming Factors 

With your data stored as factors you can use the `plot()` function to quickly see the number of observations in each factor level

```{r plot_factor}
plot(data$HighestCare)
```
However say we wanted to rename the Negative group from Neg to Negative.

```{r rename_factor}

levels(data$HighestCare)[1] <- "Negative"
plot(data$HighestCare)
```
We can also use this to rename any NA levels we may have. For example, lets make the `Ever.On.Ventilator` column a factor and plot it.

```{r rename}
data$Ever.On.Ventilator <- factor(data$Ever.On.Ventilator)
plot(data$Ever.On.Ventilator)

# But what about the NAs samples that we know are there?

# turn the missing values into a factor levels with the addNA() function
data$Ever.On.Ventilator <- addNA(data$Ever.On.Ventilator)
levels(data$Ever.On.Ventilator)

# And rename the NA
levels(data$Ever.On.Ventilator)[3] <- "Not Applicable"

# now we can replot and see this new added level
plot(data$Ever.On.Ventilator)

```
Challenge!
```{r challenge_7}
# Now that we have renamed the Ever.On.Ventilator column, can you reorder the levels so that the "Not Applicable" patients are plotted before  the "N" patients?
data$Ever.On.Ventilator <- factor(data$Ever.On.Ventilator, levels = c("Not Applicable","N","Y"))

plot(data$Ever.On.Ventilator)
```
##### Saving and exporting data

Now that we have done a variety of data manipulations, we want to save our output.

```{r write_day}

# lets output in our results folder
write.csv(data, file="/Users/ewitkop/Library/CloudStorage/Box-Box/EW_Bioinformatics_Postdoc_Research/ADMINISTRATIVE/Data_Carpentry_Workshop_June_2022/WORKSHOP_CODE/Data_Carpentry_Workshop_2022/results/data_day1.csv", row.names = FALSE)


```


